GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-12-08 writeUps/paper.tex
filled in missing citations/refs

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance'' \citealt{}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Then, the standard error matrix for $[\bm{\hat{\alpha}},\bm{\hat{\beta}}$ can be estimated as \citep[][ch. 7]{boosStefanskiBook}:
\begin{equation}\label{eq:sandwich}
  \widehat{var}(\bm{\hat{\beta}})=A^{-1}BA'
\end{equation}
Where
\begin{equation}\label{eq:Amat}
  A=\sum_i\frac{\partial}{\partial [\alpha,\beta]'} [Z_i\Omega_i,\Psi_i]'
\end{equation}
and
\begin{equation}\label{eq:Bmat}
  B=\sum_i [Z_i\Omega_i, \Psi_i]'[Z_i\Omega_i, \Psi_i]
\end{equation}
where $[Z_i \Omega_i, \Psi_i]$ is the stacked vector of estimating equations evaluated at $Z_i$, $\bxsi$, $S_i$, and $Y_i$.
The appendix gives details of this calculation, giving a consistent variance estimator.

@@

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEst}
  \begin{split}
    \hat{\eff0}_{int}&\equiv \hat{\beta}_3+\bm{\beta}_6'\overline{\bxy}_{Z=1,S=0}\\
    \hat{\eff1}_{int}&\equiv \hat{\beta}_3+(\bm{\hat{\beta}_6}+\bm{\hat{\beta}_7})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\hat{\eff0}_{G}$ and $\hat{\eff1}_{int}$ are M-estimators, with $\hat{\eff0}_{int}\rightarrow_p\eff0$ and $\hat{\eff1}_{G}\rightarrow_p\eff1$ as $n\rightarrow\infty$.
  Under suitible regularity conditions, they are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

@@

Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatterplots, showing estimation error for GEEPERs, the mixture estimator, and the PSW estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:rmse} shows root mean squared error (RMSE) estimates under a somewhat wider set of circumstances when $\muc1=0.3$.
The leftmost panel of Figure \ref{fig:boxplots} and the top line of Tables \ref{tab:rmse} and \ref{tab:coverage} show results under ``standard'' conditions---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
In this case, GEEPERs is roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it is much noisier, and when $\alpha=0.5$ it is only slightly less precise.
The mixture estimator's bias depends on $\muc1$---when $\muc1=0$, the bias is slight, and when $\muc1=0.3$ the bias is more substantial. This is somewhat surprising, because when $\muc1=0$, there is no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
The PSW estimator is also biased---due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:pi}--inherent in the data generating model. On the other hand, the PSW estimator has easily the lowest sampling variance of the three estimators.
In fact, Table \ref{tab:rmse} shows that despite its bias, the PSW estimator has lower RMSE than GEEPERs when $\alpha=0$ or 0.2.
The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:rmse} shows a similar set of results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
In this case, GEEPERs and PSW estimator behave in a similiar fashion to the normal case, but the mixture estimator shows a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.

@@

In a recent educational field experiment, middle school students completing their math schoolwork on a computer program were randomized between four conditions. In all four conditions, students worked on their assigned math program in class on 12 different occasions---a prior assessment, a mid-test, a post-test, and nine learning sessions---spaced throughout the school year.  In three of the conditions, during the learning sessions, students worked through a series of math problems, while the forth condition was a computer game designed to teach math concepts. In this illustration, we will consider three different treatment effects, comparing one of the four conditions (which we arbitrarily label ``treatment''), with each of the three others (Business as Usual, or ``BAU'', From-Here-To-There!, or ``FH2T,'' and DragonBox), separately. During learning sessions, students in the treatment condition received immediate, informative error messages after entering a wrong answer, and were required to enter a correct answer in order to proceed. They also had access to a series of hints for each problem, the last of which (called the ``bottom out'' hint) contained the correct answer. Students in the other three conditions did not have access bottom out hints; our illustration will focus on that software feature.

The literature on hints in online learning is mixed \citep{stuff}, in particular when it comes to requesting a bottom-out hint (called ``bottoming out''). On the one hand, hints that include the correct answer are essentially worked examples, which can be beneficial to learning \citep{workedexamplestuff}. On the other hand, they allow students to ``game the system,'' by simply requesting enough hints to see the right answer, and use the software without ever working to solve any problems. However, in this experiment, the ability to bottom out was only one of several differnces between the treatment conditions. One way to disentangle the role of bottoming out in any treatment effect would be to estimate average treatment effects separately for subjects who would (if assigned to treatment) bottom out, and for subjects who would not. If hints (that do not supply the correct answer) and error messages are helpful, but the ability to bottom out is harmful to learning, then we might expect to estimate a positive effect of being assigned to the feedback condition for students who do not bottom out, but a negative effect for students who do. If hints are helpful for learning only if they can be read as worked examples, then we might expect no effect of condition on students who do not bottom out, but a positive effect for students who do. (These hypothetical results may be masked by other treatment effect heterogeneity between students who would or wouldn't bottom out, but the results could still be informative.)

@@

The model was fairly successful in distinguishing bottom-outers from non-bottom-outers---the AUC, evaluated using out-of-sample predictions in a 10-fold cross-validation, was roughly 0.795. That is, in roughly 80\% of bottom-outer/non-bottom-outer pairs of subjects, the bottom-outer will have a higher predicted probability from the model.
Table \ref{tab:results} show the model's coefficient estimates. Notably, conditional on the other covariates, students with higher fifth-grade standardized test scores were less likely to request many bottom out hints, as were students who spent little time on task during the pretest, while IEP and ESOL students, and students without baseline measurement data, were relatively unlikely to be bottom-outers. The relationship between perceptual sensitivity and bottom out requests was not monotonic---students with average levels were more likely than those with more extreme high or low levels to be bottom-outers; bottom-outers tended to have lower math self-efficacy, but the data are also consistent with a small positive relationship, and there appears to be little to no relationship between math anxiety and bottom-outer status at all.

@@

replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations:
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\Psi_i=\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

@@

\subsection{Proof for Proposition \ref{prop:interactions}}
First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ due to \eqref{ass:rand}}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}

implying that
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\bm{\beta_1}\bxy+\beta_2\pp+\bm{\beta_5}'\bxy\pp\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_3+(\bm{\beta_1}+\bm{\beta_6})\bxy+(\beta_3+\beta_4)S+(\bm{\beta_5}+\bm{\beta_7})'\bxy S
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]&=\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[\pp]+\bm{\beta_5}'\EE[\bxy\pp]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[S]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy S]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[R]+\bm{\beta_5}'\EE[\bxy R]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[R]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy R]\right\}\\
\end{align*}
Analogous reasoning leads to expressions for $\EE[RY]$, $\EE[\bxy Y]$, $\EE[ZY]$,  $\EE[ZRY]$, $\EE[\bxy YZ]$, and $\EE[\bxy RZY]$.
These, in turn, give rise to estimating equations
\begin{equation*}
  \displaystyle\sum_i \begin{pmatrix}
    Y_i-\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    R_iY_i-\beta_0R_i-\bm{\beta_1}\bxy_i R_i-\beta_2R_i^2-\bm{\beta_5}'\bxy_iR_i^2-Z_i\left\{\beta_3R_i-\bm{\beta_4}R_i^2-\bm{\beta_6}'\bxy_iR_i-\bm{\beta_7}'\bxy_i R_i^2\right\}\\
    \bxy_i Y_i-\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    Z_iY_i-Z_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iR_iY_i-Z_iR_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iY_i\bxy_i-Z_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    R_iZ_iY_i\bxy_i-Z_iR_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}
  \end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\\0\\0\\0\end{pmatrix}
\end{equation*}
These are the estimating equations for the regression model \eqref{eq:fullInteractionModel}, fit by OLS. Now note that
\begin{equation*}
  \EE[Y_T-Y_C|\st=0]=\beta_0^T-\beta_0^C+(\bm{\beta_1}^{T'}-\bm{\beta_1}^{C'})\EE[\bxy|S=0]=\beta_3+\beta_6'E[\bxy|S=0]
\end{equation*}
and
\begin{equation*}
  \EE[Y_T-Y_C|\st=1]=\beta_0^T+\beta_2^T-(\beta_0^C+\beta_2^C)+(\bm{\beta_1}^{T'}+\bm{\beta_3^{Ti}}-\bm{\beta_1}^{C'}-\bm{\beta_3^{C'}})\EE[\bxy|S=0]=\beta_3+\beta_4+(\beta_6'+\beta_7')E[\bxy|S=1]
\end{equation*}
the result follows.

@@

So we're estimating principal effects in two stages.
First, fit the model
\begin{equation}\label{eq:psMod}
  Pr(S_{Ti}=1|\bm{\tilde{w}}_i)=f(\bm{\alpha}'\bm{\tilde{w}}_i)
\end{equation}
for some inverse link function $f(\cdot)$, where $\bm{\tilde{x}}_i=[1,\bm{w}_i]$, using (observed) values from the treatment group, and estimating $\hat{\alpha}$.
Then let
\begin{equation}\label{eq:ps}
  \hat{p}_i=f(\bm{\hat{\alpha}}'\bm{\tilde{w}}_i)
\end{equation}
for all subjects in the experiment.

>>>>

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance'' \citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Then, the standard error matrix for $[\bm{\hat{\alpha}},\bm{\hat{\beta}}$ can be estimated as \citep[][ch. 7]{boosStefanskiBook}:
\begin{equation}\label{eq:sandwich}
  \widehat{var}(\bm{\hat{\beta}})=A^{-1}BA'
\end{equation}
Where
\begin{equation*}
  A=\sum_i\frac{\partial}{\partial [\alpha,\beta]'} [Z_i\Omega_i,\Psi_i]'
\end{equation*}
and
\begin{equation*}%\label{eq:Bmat}
  B=\sum_i [Z_i\Omega_i, \Psi_i]'[Z_i\Omega_i, \Psi_i]
\end{equation*}
where $[Z_i \Omega_i, \Psi_i]$ is the stacked vector of estimating equations evaluated at $Z_i$, $\bxsi$, $S_i$, and $Y_i$.
The appendix gives details of this calculation, giving a consistent variance estimator.

@@

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEst}
  \begin{split}
    \hat{\eff0}_{int}&\equiv \hat{\beta}_2+\bm{\hat{\gamma}_3}'\overline{\bxy}_{Z=1,S=0}\\
    \hat{\eff1}_{int}&\equiv \hat{\beta}_2+\hat{\beta}_3+(\bm{\hat{\gamma}_3}+\bm{\hat{\gamma}_4})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\hat{\eff0}_{G}$ and $\hat{\eff1}_{int}$ are M-estimators, with $\hat{\eff0}_{int}\rightarrow_p\eff0$ and $\hat{\eff1}_{G}\rightarrow_p\eff1$ as $n\rightarrow\infty$.
  Under suitible regularity conditions, they are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

@@

Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatterplots, showing estimation error for GEEPERs, the mixture estimator, and the PSW estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:rmse} shows root mean squared error (RMSE) estimates under a somewhat wider set of circumstances when $\muc1=0.3$.
The leftmost panel of Figure \ref{fig:boxplots} and the top line of Tables \ref{tab:rmse} and \ref{tab:coverage} show results under ``standard'' conditions---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
In this case, GEEPERs is roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it is much noisier, and when $\alpha=0.5$ it is only slightly less precise.
The mixture estimator's bias depends on $\muc1$---when $\muc1=0$, the bias is slight, and when $\muc1=0.3$ the bias is more substantial. This is somewhat surprising, because when $\muc1=0$, there is no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
The PSW estimator is also biased---due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:PI}--inherent in the data generating model. On the other hand, the PSW estimator has easily the lowest sampling variance of the three estimators.
In fact, Table \ref{tab:rmse} shows that despite its bias, the PSW estimator has lower RMSE than GEEPERs when $\alpha=0$ or 0.2.
The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:rmse} shows a similar set of results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
In this case, GEEPERs and PSW estimator behave in a similiar fashion to the normal case, but the mixture estimator shows a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.

@@

In a recent educational field experiment, middle school students completing their math schoolwork on a computer program were randomized between four conditions. In all four conditions, students worked on their assigned math program in class on 12 different occasions---a prior assessment, a mid-test, a post-test, and nine learning sessions---spaced throughout the school year.  
The four conditions ranged from a game---DragonBox---to the active control condition in which students simply used the computer to work on a sequence of math problems. 
In the DragonBox game, students learned to solve algebraic equations by isolating a box containing a dragon; as students progressed in the game, pictures of monsters were replaced by mathematical symbols. 
Another condition, the From Here to There (FH2T) program, used visual cues such as color and spacing to teach algebraic concepts, and allowed students to solve equations by manipulating expressions via dragging and dropping.
In the ASSISTments condition, students used computers to work through a series of math problems taken from the district's textbook. While working, students had access to a series hints for each problem, culminating in a ``bottom out'' hint that contained the answer. Students in the ASSISTments condition were given immediate feedback on their answers and had to enter the correct answer in order to proceed to the next problem. 
In the active control condition, labeled ``business as usual'' (BAU), the computer work mimicked traditional pencil-and-paper mathematics coursework---students worked through a series of math problems without access to hints and received feedback only after completing the assignment. 
In this illustration, we will compare the ASSISTments condition (which we arbitrarily label ``treatment'')---the only condition featuring bottom-out hints---with each of the three others separately. Our goal is to estimate different treatment effects for groups of students who would, if assigned to the ASSISTments condition, request few or many bottom out hints.

The literature on hints in online learning is mixed \citep[see, e.g.,][]{aleven2016help,goldin2012learner,sales2021student} in particular when it comes to requesting a bottom-out hint (called ``bottoming out''). On the one hand, hints that include the correct answer are essentially worked examples, which can be beneficial to learning \citep[e.g.]{sweller1985use}. On the other hand, they allow students to ``game the system,'' by simply requesting enough hints to see the right answer, and use the software without ever working to solve any problems \citep[e.g.][]{guo2008trying}. However, in this experiment, the ability to bottom out was only one of several differences between the treatment conditions. One way to disentangle the role of bottoming out in any treatment effect would be to estimate average treatment effects separately for subjects who would (if assigned to treatment) bottom out, and for subjects who would not. If hints (that do not supply the correct answer) and error messages are helpful, but the ability to bottom out is harmful to learning, then we might expect to estimate a positive effect of being assigned to the feedback condition for students who do not bottom out, but a negative effect for students who do. If hints are helpful for learning only if they can be read as worked examples, then we might expect no effect of condition on students who do not bottom out, but a positive effect for students who do. (These hypothetical results may be masked by other treatment effect heterogeneity between students who would or wouldn't bottom out, but the results could still be informative.)

@@

The model was fairly successful in distinguishing bottom-outers from non-bottom-outers---the AUC, evaluated using out-of-sample predictions in a 10-fold cross-validation, was roughly 0.795. That is, in roughly 80\% of bottom-outer/non-bottom-outer pairs of subjects, the bottom-outer will have a higher predicted probability from the model.
Table \ref{fig:psMod} show the model's coefficient estimates. Notably, conditional on the other covariates, students with higher fifth-grade standardized test scores were less likely to request many bottom out hints, as were students who spent little time on task during the pretest, while IEP and ESOL students, and students without baseline measurement data, were relatively unlikely to be bottom-outers. The relationship between perceptual sensitivity and bottom out requests was not monotonic---students with average levels were more likely than those with more extreme high or low levels to be bottom-outers; bottom-outers tended to have lower math self-efficacy, but the data are also consistent with a small positive relationship, and there appears to be little to no relationship between math anxiety and bottom-outer status at all.

@@

Replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations $\sum_{i=1}^{n_C}\Psi_i=\bm{0}$ or 
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

@@

\subsection{Proof for Proposition \ref{prop:interactions}}
First of all, by \eqref{eq:interaction}, 
\begin{equation*}
  \EE[Y_T-Y_C|\st=0]=\EE[Y|Z=1,\st=0]-\EE[Y|Z=1,\st=0]=\beta_2+\bm{\gamma_3}'\EE[\bxy|\st=0]
\end{equation*}
and
\begin{equation*}
  \EE[Y_T-Y_C|\st=1]=\EE[Y|Z=1,\st=1]-\EE[Y|Z=1,\st=1]=\beta_2+\beta_3+(\bm{\gamma_3}'+\bm{\gamma_4}')\EE[\bxy|\st=1]
\end{equation*}
Furthermore, $\overline{\bxy}_{Z=1,S=0}\rightarrow \EE[\bxy|\st=0]$ and $\overline{\bxy}_{Z=1,S=1}\rightarrow \EE[\bxy|\st=1]$ as $n\rightarrow \infty$.

We will show that the estimated coefficients from model \eqref{eq:interaction}, but with $R$ replacing $\st$, fit with OLS, are consistent for $\bm{\beta}$ and $\bm{\gamma}$ from \eqref{eq:interaction}.

First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ (due to randomization)}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}
implying that, according to \eqref{eq:interaction},
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\beta_1\pp+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}\pp\\
                  &=\beta_0+\beta_1R+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}R\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_2+(\beta_1+\beta_3)S+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy S\\
    &=\beta_0+\beta_2+(\beta_1+\beta_3)R+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy R
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]&=\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  &=\beta_0+\beta_1\EE[R]+\beta_2\EE[Z]+\beta_3\EE[ZR]+\bm{\gamma_1}'\EE[\bxy]+\bm{\gamma_2}'\EE[\bxy R]+\bm{\gamma_3}'\EE[\bxy Z]+\bm{\gamma_4}\EE[\bxy RZ]
\end{align*}

Analogous reasoning leads to expressions for $\EE[RY]$, $\EE[\bxy Y]$, $\EE[ZY]$,  $\EE[ZRY]$, $\EE[\bxy YZ]$, and $\EE[\bxy RZY]$.
These, in turn, give rise to estimating equations
\begin{equation*}
  \displaystyle\sum_i \begin{pmatrix}
  Y_i\\
  \phantom{}\\
  Y_i\ri\\
    \phantom{}\\
  Y_iZ_i\\
  \phantom{}\\
  Y_iZ_i\ri\\
  \phantom{}\\
  Y_i\bxy_i\\
  \phantom{}\\
  Y_i\bxy_i\ri\\
  \phantom{}\\
  Y_i\bxy_iZ_i\\
  \phantom{}\\
  Y_i\bxy_iZ_i\ri\\
  \phantom{}\end{pmatrix} - \begin{pmatrix}
    \beta_0+\beta_1R_i+\beta_2Z_i+\beta_3Z_iR_i\phantom{-----}\\
    \hfill+\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    \beta_0\ri+\beta_1R_i^2+\beta_2Z_i\ri+\beta_3Z_iR_i^2\phantom{-----}\\
    \hfill+\ri\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    (\beta_0+\beta_2)Z_i+(\beta_1+\beta_3)R_iZ_i\phantom{-----}\\
    \hfill+Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    \beta_0Z_i\ri+(\beta_1+\beta_3)Z_iR_i^2+\beta_2Z_i\ri\phantom{-----}\\
    \hfill+Z_iR_i\left\{\bm{\gamma_1}'+\bm{\gamma_3}'+R_i(\bm{\gamma_2}'+\bm{\gamma_4}')\right\}\bxy \hfill\\
    \beta_0\bxyt+\beta_1R_i\bxyt+\beta_2Z_i\bxyt+\beta_3Z_iR_i\bxyt\phantom{---}\\
    \hfill {}+\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
    \beta_0\ri\bxyt+\beta_1R_i^2\bxyt+\beta_2Z_i\ri\bxyt+\beta_3Z_iR_i^2\bxyt\phantom{---}\\
    \hfill {}+\ri\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
        (\beta_0+\beta_2)Z_i\bxyt+(\beta_1+\beta_3)R_iZ_i\bxyt\phantom{-----}\\
    \hfill {}+Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
\beta_0\ri Z_i\bxyt+\beta_1R_i^2Z_i\bxyt+\beta_2\ri Z_i\bxyt+\beta_3Z_iR_i^2\bxyt\phantom{---}\\
    \hfill {}+\ri Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
  \end{pmatrix}=\bm{0}
\end{equation*}
These are the estimating equations for the regression model \eqref{eq:interaction}, with $R$ replacing $\st$, fit by OLS. 
Under standard OLS regularity conditions, the estimated coefficients of that model are consistent, completing the proof.

@@

Here we will derive the sandwich variance-covariance matrix for the GEEPERs estimate without interactions between $\bx$ and either $Z$ or $\st$---i.e. with $\bm{\gamma_2}=\bm{\gamma_3}=\bm{\gamma_4}=0$ in the notation of \eqref{eq:interaction}---and estimating principal scores using a generalized linear model.

We propose estimating principal effects in two stages.
First, fit the model
\begin{equation}\label{eq:psMod}
  \ppi=Pr(\sti=1|\bxsi)=f(\bm{\alpha}'\bxsit)
\end{equation}
for some inverse link function $f(\cdot)$, where $\bxsit=[1,\bxsi]$, using (observed) values from the treatment group, and estimating $\hat{\alpha}$.
Then let
\begin{equation}\label{eq:ps}
  \hat{p}_i=f(\bm{\hat{\alpha}}'\bxsit)
\end{equation}
for all subjects in the experiment.

Finally, fit model
\begin{equation}\label{eq:regression}
  Y_i=\beta_0+\beta_1r_i+\beta_2Z_i+\beta_3Z_i\ri+\bm{\gamma}'\bxy_i+\epsilon_i
\end{equation}
to estimate $\bm{\beta}$ and hence principal effects, where $\bm{x}_i$ is a set of covariates predictive of $Y$ within principal strata.
Let $\bm{\beta}=[\beta_0,\beta_1,\beta_2,\beta_3,\bm{\gamma}']'$.
